Impact of transmission type on Miles per Gallon
==============================================================

Executive Summary
-----------------

In this analysis MPG and relationship with automatic vs. manual transmission have been compared for 32 different car models. Results obtained can be summarized in 6 key points:
- cars with **manual transmission are better** for **MPG**; while cars with automatic transmission have a lower level of MPG
- holding other features constant, **manual transmission** leads to a positive **increase in MPG of about 5.2**
- **Horsepower** is an **important variable** when assessing MPG jointly with transmission type: **the more the horsepower the less the MPG**
- **Results** obtained **are robust** and subject to relatively **little uncertainity**
- Model obtained suffers anyway from the fact that **few observations are available for cars with hp > 200 and manual transmission**
- **Taking more observations** for cars with hp > 200 and manual transmission could help **strengthen results in the hp > 200 range**


Data exploration
----------------
All variables in the mtcars data frame are store in **numeric format**. However cyl, vs, am, gear and carb will be treated as factor variables. In particular for transmission type it's better to also redefine levels in a readable format:
```{r echo=FALSE}
data <- mtcars
data$cyl <- as.factor(data$cyl)
data$vs <- as.factor(data$vs)
data$gear <- as.factor(data$gear)
data$carb <- as.factor(data$carb)
```

```{r}
data$am <- as.factor(data$am)
levels(data$am) <- c("automatic","manual")
```

Doing so we can immediately compare MPG to transmission type **(see Fig. 1 in the appendix)**. From this chart is clearly visible that **cars with manual transmission are better for MPG**.

Linear model fitting
--------------------

The strategy applied for selecting the best model is 3-step and aims at preserving **am** as predictor so to answer the relevant question.

**Step 1:** fit a uni-variate model with **am** as predictor
```{r}
vars <- "am"
outcome <- "mpg"
formula <- paste0(outcome,"~",paste0(vars,collapse="+"))
fit <- lm(formula,data)
```

**Step 2:** keeping **am** as predictor, **generate all possible bivariate models** with remaining columns. Compare bi-variate models with uni-variate model using **anova method** and keep the one whose comparison has the **smallest p-value**, but only if **am** is statistically significant.

**Step 3:** Repeat step 2 continuously adding the best predictor until p-value of **am** predictor looses statistic significance

```{r}
#step 2 and 3
loop <-  TRUE
while(loop == TRUE){
        #generate all possible formulas, break if all vars are being used
        remainingVars <- setdiff(colnames(data),c(outcome,vars))
        if(length(remainingVars) == 0){break}
        newFormulas <- paste0(formula,"+",remainingVars)
        
        #generate new linear models
        numModels <- length(newFormulas)
        newFit <- lapply(1:numModels, function(x) lm(newFormulas[x],data))
        
        #compare new linear models with baseline model
        anovaPvalues <- sapply(1:numModels, function(x) anova(fit,newFit[[x]])[2,6])
        candidateFit <- which.min(anovaPvalues)
        
        #update baseline model if a new better model has been found
        betterModel <- anovaPvalues[candidateFit] < 0.05
        amSignif <- summary(newFit[[candidateFit]])$coefficients[[2,4]] < 0.05
        if (betterModel & amSignif){
                formula <- newFormulas[candidateFit]
                fit <- newFit[[candidateFit]]
                vars <- c(vars,remainingVars[candidateFit])
        } else {loop <- FALSE}        
}
```

The final output is a bi-variate model with **am** and **hp** as predictors. This result makes sense also at an intuitive level since **horsepower** is something to keep into consideration when evaluating fuel consumption of engines.

************
Selected model assessment
-------------------------
Let's check how good is the model that was found:
```{r echo=FALSE}
summary(fit)
```

All coefficients are statistically highly significant. Looking at the **residual diagnostic plots (see Fig. 2)** we can see that they are normally distributed with no visible pattern.

Conclusions and uncertainty in results
----------------------
We can conclude that: 

- cars with manual transmission are better for **MPG**
- according to our model a hypothetical **car with automatic transmission and 50 hp** has an estimated **MPG of `r (fit$coefficients[3]*50)+fit$coefficients[1]`**:
```{r eval=FALSE}
fit$coefficients[3]*50+fit$coefficients[1]
```

- holding horsepower constants choice of **manual transmission allows to travel `r fit$coefficients[2]`** more per each gallon
- horsepower has also an effect on MPG and, for the same kind of transmission, **each horsepower added, decreases MPG by `r -fit$coefficients[3]`**


We can see that all obtained coefficients have relatively narrow 95% confidence intervals:
```{r echo=FALSE}
confint(fit)
```

Lastly, we can check if any point badly influences the results. Actually 2 car models are **highly influential (see Fig. 3)**:
```{r echo=FALSE}
influentialPoints <- apply(influence.measures(fit)$is.inf,1,sum)>0
row.names(data[influentialPoints,])
```

To assess the impact of influential points we can fit a new model excluding these observations and adding an interaction term between hp and transmission type: **mpg~am+hp+am*hp**. P-values and residuals plots **(see Fig. 4)** indicate this model is valid.
```{r echo=FALSE}
newdata <- data[!influentialPoints,]
notInfluencedfit <- lm(mpg~am+hp+am*hp,newdata)
summary(notInfluencedfit)$coef[,4]
```
Looking at fitted lines **(see Fig. 5)**, we now see that **better performance of manual transmission is now not true for values of hp higher than 200**. To decrease the uncertainty on transmission performance in the range of 200 hp, **more observations** of cars with manual transmission and hp in this range could be taken.

************
Appendix
----------------

**Figure 1: Distribution of MPG by tyoe if transmission**

```{r echo=FALSE}
boxplot(data$mpg~data$am,col="blue",xlab="Transmission type", ylab="MPG")
```

**Figure 2: residuals diagnostics plots for selected linear model**

```{r echo=FALSE,fig.width=9.5,fig.height=9.5}
par(mfrow=c(2,2))
plot(fit)
```
************

**Figure 3: regression lines and influential points**

```{r echo=FALSE}
plot(data$hp,data$mpg,col=data$am,pch=19, xlab="Horsepower",ylab="MPG")
legend(260,33,c("manual","automatic"), lty = c(1,1), col = c("red","black"))
abline(fit$coefficients[1],fit$coefficients[3])
abline(fit$coefficients[1]+fit$coefficients[2],fit$coefficients[3],col="red")
text(data$hp[influentialPoints],data$mpg[influentialPoints],rownames(data)[influentialPoints],pos=2)
```


**Figure 4: residuals diagnostics plots for linear model without influential points**

```{r echo=FALSE,fig.width=10,fig.height=10}
par(mfrow=c(2,2))
plot(notInfluencedfit)
```
************

**Figure 5:regression lines for model without influential points and interaction (hp,am)**

```{r echo=FALSE}
plot(newdata$hp,newdata$mpg,col=newdata$am,pch=19, xlab="Horsepower",ylab="MPG")
legend(190,33,c("manual","automatic"), lty = c(1,1), col = c("red","black"))
abline(notInfluencedfit$coefficients[1],notInfluencedfit$coefficients[3])
abline(notInfluencedfit$coefficients[1]+notInfluencedfit$coefficients[2],notInfluencedfit$coefficients[3]+notInfluencedfit$coefficients[4],col="red")
```